Live Functional Programming with Typed Holes

### 2018-07-10 examples-1.tex
tweaks to the figures and text

Let us now consider a second more sophisticated example: an incomplete implementation of the recursive quicksort function, shown in Fig.~\ref{fig:qsort-example-code}. So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid during class) has filled in the base case, and in the recursive case, has selected the head of the list as the pivot, partitioned the remainder of the list relative to the pivot and made the two recursive calls. A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables.

At the bottom of the cell in Fig.~\ref{fig:qsort-example-code}, the programmer has applied \li{qsort} to an example list. However, 
the indeterminate result of this function application appears rather uninteresting: it is simply an instance of hole \li{1}, which serves only to confirm that evaluation went through the recursive case of \li{qsort}. 
More interesting is the live context inspector, shown in three states in Fig.~\ref{fig:qsort-sidebars}, which provides feedback about the values of the variables in scope at hole \li{1} from the the various instances of hole \li{1} that appear in the result. For example, the initial state of the context inspector (Fig.~\ref{fig:qsort-sidebars}, left) shows the closure at the instance of hole \li{1} that appears immediately in the result due to the outermost application of \li{qsort}. From this, the programmer can confirm (or the lecturer can visually point out) that 
the lists \li{smaller} and \li{bigger} computed by the call to \li{partition} are appropriately named, and observe that they are not yet themselves sorted\todo{are there studies to cite here about how concrete examples help people in programming? math?}.

The results from the subsequent recursive calls, \li{r_smaller} and \li{r_bigger}, are again hole instances, \li{1:2} and \li{1:3}. 
The programmer can click on either of these hole instances to reveal the associated closures from the corresponding recursive calls. 
For example, clicking on \li{1:3} reveals the hole closure from the \li{r_bigger} recursive call as shown in Fig.~\ref{fig:qsort-sidebars} (middle). 
From there, the programmer can click another hole closure, e.g. \li{1:6} to reveal the hole closure from the subsequent \li{r_smaller} recursive call as shown in Fig.~\ref{fig:qsort-sidebars} (right). 
Notice that the path from the root to the selected hole closure is reported as shown at the bottom of the context inspector in Fig.~\ref{fig:qsort-sidebars}. 
In exploring these paths rooted at the result, the programmer can develop concrete intuitions about the recursive structure of the computation (e.g. by following through to the base case as shown in Fig.~\ref{fig:qsort-sidebars}) even before the program is complete. 
% (We leave to future work the pedagogical question of whether such early feedback might help students understand recursion.)

>>>>

Let us now consider a second more sophisticated example: an incomplete implementation of the recursive quicksort function, shown in Fig.~\ref{fig:qsort-example-code}. So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid during class) has filled in the base case, and in the recursive case, has selected the head of the list as the pivot, partitioned the remainder of the list relative to the pivot and made the two recursive calls. A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables to fill the hole with an appropriate list expression, as indicated by the type inspector in Fig.~\ref{fig:qsort-type-inspector}.

At the bottom of the cell in Fig.~\ref{fig:qsort-example-code}, the programmer has applied \li{qsort} to an example list. However, 
the indeterminate result of this function application is simply an instance of hole \li{1}, which serves only to confirm that evaluation went through the recursive case of \li{qsort}. 
More interesting is the live context inspector, shown in three states in Fig.~\ref{fig:qsort-sidebars}, which provides feedback about the values of the variables in scope at hole \li{1} from the the various instances of hole \li{1} that appear in the result. For example, in its initial state (Fig.~\ref{fig:qsort-sidebars}, left) it shows the closure at the instance of hole \li{1} that appears immediately in the result due to the outermost application of \li{qsort}. From this, the programmer can confirm (or the lecturer can visually point out) that 
the lists \li{smaller} and \li{bigger} computed by the call to \li{partition} are appropriately named, and observe that they are not yet themselves sorted\todo{are there studies to cite here about how concrete examples help people in programming? math?}.

The results from the subsequent recursive calls, \li{r_smaller} and \li{r_bigger}, are again hole instances, \li{1:2} and \li{1:3}. 
The programmer can click on either of these hole instances to reveal the associated closures from the corresponding recursive calls. 
For example, clicking on \li{1:3} reveals the hole closure from the \li{r_bigger} recursive call as shown in Fig.~\ref{fig:qsort-sidebars} (middle). 
From there, the programmer can click another hole closure, e.g. \li{1:6} to reveal the hole closure from the subsequent \li{r_smaller} recursive call as shown in Fig.~\ref{fig:qsort-sidebars} (right). 
Notice in each case that the path from the result to the selected hole closure is reported as shown at the bottom of the context inspector in Fig.~\ref{fig:qsort-sidebars}. 
In exploring these paths rooted at the result, the programmer can develop concrete intuitions about the recursive structure of the computation (e.g. by following through to the base case as shown in Fig.~\ref{fig:qsort-sidebars}) even before the program is complete. 
% (We leave to future work the pedagogical question of whether such early feedback might help students understand recursion.)

### 2018-07-10 qsort-example.tex
tweaks to the figures and text

\begin{figure}[t]
\begin{subfigure}[t]{0.65\textwidth}
\centering
\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/qsort-code.png}
\caption{The result of evaluation is a hole closure.}
\label{fig:qsort-example-code}
\end{subfigure}
~
\begin{subfigure}[t]{0.33\textwidth}
\centering
\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/type-inspector.png}
\vspace{-2px}
\caption{The type inspector
continuously provides static information about the expression at the cursor. Here, it tells us that the hole at the cursor should be filled by a list expression. Holes have hole (i.e. unknown) type.
}
\label{fig:qsort-type-inspector}
\end{subfigure}

\vspace{10px}

@@

\vspace{-6px}
\end{figure}

>>>>

\begin{figure}[t]
\begin{subfigure}[t]{0.70\textwidth}
\centering
\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/qsort-code.png}
\vspace{-3px}
\caption{The result of evaluation is a hole closure.}
\label{fig:qsort-example-code}
\end{subfigure}
~
\begin{subfigure}[t]{0.29\textwidth}
\centering
\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/type-inspector.png}
\vspace{-3px}
\caption{The type inspector provides static information about the term at the cursor. Here, it tells us that the hole should be filled by a list expression. Holes have hole (i.e. unknown) type, which  is universally consistent (see Sec.~\ref{sec:calculus} and \cite{popl-paper}).
}
\label{fig:qsort-type-inspector}
\end{subfigure}

\vspace{8px}

@@

\vspace{-8px}
\end{figure}
